Liveness-Based Garbage Collection for Lazy Languages

### 2015-11-27 paper/paper.tex
fixed minefield semantics
H and l fixed

\begin{center}\footnotesize
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\
\hline
\hline 
          & $\rho,\, (\rho', \ell, e)\!:\!S,\, H,\, \kappa
  \rightsquigarrow \rho',\, S,\, H[\ell := \kappa],\, e$    &  \sc{const}
\\
\hline
          & {$\rho, \,(\rho', \ell, e)\!:\!S,\, H,\, (\CONS~x~y)
\rightsquigarrow
$  $\rho',\, S,\, H[\ell := (\rho(x), \rho(y))],\, e$}     &  \sc{cons} \\
\hline

$H(\rho(x)) \mbox{ is } (v, d)$ & $\rho,\, (\rho', \ell, e)\!:\!S,\, H,\,
(\CAR~x)  \rightsquigarrow \rho',\, S,\, H[\ell := v],\, e$      &
\sc{car-select} \\
\hline

$H(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$ & $\rho,\, S,\,  H,\,
(\CAR~x)  \rightsquigarrow \rho', \,(\rho, addr(\langle s, \rho'\rangle), (\CAR~x))\!:\!S,\, H,\, s$      &
\sc{car-1-clo} \\
\hline

$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho,\, S,\, H,\, (\CAR~x)
\rightsquigarrow
\rho',\, (\rho, \rho(x), (\CAR~x))\!:\!S, \,H,\, s$      &
\sc{car-clo}
\\
\hline

$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
 & {$\rho,\, (\rho', \ell, e)\!:\!S,\, H,\, (+~x~y)  \rightsquigarrow$
$\rho', \,S,\, H[\ell := H(\rho'(x)) + H(\rho'(y))],\, e$}      &
\sc{prim-add} \\
\hline

$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho,\, S,\, H,\, (+~x~y)
\rightsquigarrow
\rho',\, (\rho, \rho(x), (+~x~y))\!:\!S,\, H,\, s$      &
\sc{prim-1-clo} \\
\hline
$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & $\rho,\, S,\, H,\, (+~x~y)
\rightsquigarrow
\rho', (\rho,\, \rho(y),\, (+~x~y))\!:\!S,\, H,\, s$      &
\sc{prim-2-clo} \\
\hline
{$\mathit{f}~\mbox{defined as}$
$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho,\, S,\, H,\,
(f~\myvec{x})  \rightsquigarrow
[\myvec{y} \mapsto \rho(\myvec{x})],\, S,\, H,\, e_{\mathit{f}}$      &
\sc{funcall} \\
\hline
$\ell$ is a new location& {$\rho,\, S,\, H,\, (\LET~x\leftarrow s~\IN~e)
  \rightsquigarrow$
$\rho\oplus[x \mapsto \ell],\, S,\, H[\ell \mapsto \langle s,
    \lfloor\rho\rfloor_{FV(s)} %% \oplus [x \mapsto  \ell] -- NO LAZY LETS
    \rangle],\, e$} &
\sc{let} \\
\hline
$H(\rho(x)) \ne 0$ & $\rho,\, S,\, H,\, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho,\, S,\, H,\,  e_1$ & \sc{if-true} \\
\hline
$H(\rho(x)) = 0$ & $\rho,\, S,\, H,\, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho,\, S, \,H, \, e_2$ & \sc{if-false} \\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho,\, S,\, H,\,
  (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho',\, (\rho, \rho(x), (\SIF~x~e_1~e_2))\!:\!S,\, H, \, s$}
&
\sc{if-clo} \\
\hline
{$H(\rho(x))~\mbox{is in WHNF with value}~v$}& $\rho,\, (\rho', \ell,
e)\!:\!S,\, H,\,
(\SRETURN~x)  \rightsquigarrow \rho',\, S,\, H[\ell := v],\, e$ &
\sc{return-whnf}\\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho,\, S,\, H,\, (\SRETURN~x)
  \rightsquigarrow$
$\rho',\, (\rho, \rho(x), (\SRETURN~x))\!:\!S,\, H,\,  s$} &
\sc{return-clo} \\
\hline
\end{tabular}
\caption{The small-step semantics for the language. \label{fig:lang-semantics}}
\end{center}
%}\comment
\end{figure*}

@@

The  initial  state  of  the  transition  system  described  above  is
$([\;]_\rho,\,   ([\;]_\rho,   \,   \rho(\ans),\,   (\print~\ans)):[\;]_{S},
[\;]_{H},\, (\mainpgm))$,  in which  $[\;]_\rho$ and $[\;]_H$  are the
empty environment and  the empty heap. The  initial stack consists
of  a single  continuation frame  in  which \ans\  is a  distinguished
variable  that  would   eventually  be  updated  with   the  value  of
(\mainpgm).  In  addition, \print\ is  a function modeling  a printing
mechanism---a   standard   run-time   support  assumption   for   lazy
languages---that prints  the value of  (\mainpgm).  The operator $:$
pushes elements on top of the stack.

@@

%% \hline
%% \makecell[t]{$GC(\rho_1, S, H_1, \sigma) = (\rho, S, H)$,}\\ $H(\rho(x)) \mbox{ is } \bot$ & $\rho, S,
%%   H, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$   &
%% \sc{car-clo-bang}
%% \\
%% \hline
%% \makecell[t]{$GC(\rho_1, S, H_1, \Lv, \sigma) = (\rho, S, H)$,}\\ $H(\rho(x))
%% \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S,
%%   H, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, x,
%%   (\CAR~x), \cred{\sigma})\!:\!S, H, s, \cred{(\clazy \cup \acar)\sigma }$         &
%% \sc{car-clo}\\
%% \hline
%% \end{tabular}

@@

%%           &\makecell{ $\rho, (\rho', x, e, \cred{\sigma'})\!:\!S,
%%   H, \kappa, \cred{\sigma}$  $\rightsquigarrow \rho', S, H[\rho'(x) :=
%%     \kappa], e, \cred{\sigma'}$ }   &  \sc{const}
%% \\
%% \hline
%%           & \makecell{$\rho, (\rho', z, e, \cred{\sigma'})\!:\!S, H, (\CONS~x~y), \cred{\sigma}$  $\rightsquigarrow
%%   \rho', S, H[\rho'(z) := (\rho(x),\rho(y))], e, \cred{\sigma'}$}     &  \sc{cons} \\
%% \hline
%% $H(\rho(x)) \mbox{ is } (l_1, l_2)$ & \makecell{$\rho, (\rho', z, e,
%%   \cred{\sigma'} )\!:\!S, H, (\CAR~x), \cred{\sigma}$  $
%%   \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e, \cred{\sigma'}$}      &
%% \sc{car-whnf} \\
%% \hline
%% $H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
%%   H, (\CAR~x), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
%%   (\CAR~x), \cred{\sigma})\!:\!S, H, s, \cred{(\clazy \cup \acar)\sigma }$}      &
%% \sc{car-clo}
%% \\

%% \hline
%% $H(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$ & $\rho,\, S,\,  H,\,
%% (\CAR~x), \cred{\sigma} \rightsquigarrow \rho', \,(\rho, addr(\langle
%% s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,\, H,\, s, \, \cred{\sigma}$      &
%% \sc{car-1-clo} \\

%% \hline
%% $H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
%%  & \makecell{$\rho, (\rho', z, e, \cred{\sigma'})\!:\!S, H,
%%   (+~x~y), \cred{\sigma}$   $\rightsquigarrow \rho', S, H[\rho'(z)
%%     := H(\rho'(x)) + H(\rho'(y))], e, \cred{\sigma'}$}      &
%% \sc{prim-whnf} \\
%% \hline
%% $H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S,
%%   H, (+~x~y), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
%%   (+~x~y), \cred{\sigma})\!:\!S, H, s, \cred{\clazy\sigma}$}      &
%% \sc{prim-1-clo} \\
%% \hline
%% $H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho,
%%   S, H, (+~x~y), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, y,
%%   (+~x~y), \cred{\sigma})\!:\!S, H, s, \cred{\clazy\sigma}$}      &
%% \sc{prim-2-clo} \\
%% \hline
%% {$\mathit{g}~\mbox{defined as}$
%% $~(\DEFINE~(g~\myvec{y})~e_{\mathit{g}})$}  & \makecell{$\rho, S, H,
%%   (g~\myvec{x}), \cred{\sigma}$  $\rightsquigarrow [\myvec{y} \mapsto
%%     \rho(\myvec{x})], S, H, e_{\mathit{g}}, \cred{\sigma}$}      &
%% \sc{funcall} \\
%% \hline
%% $\ell$ is a new location& \makecell{$\rho, S, H, (\LET~x\leftarrow
%%   s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
%%   \ell]\rangle], e, \cred{\sigma}$} &
%% \sc{let} \\ 
%% \hline
%% $H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
%%   \cred{\sigma}$  $\rightsquigarrow \rho, S, H,  e_1, \cred{\sigma}$} & \sc{if-true} \\
%% \hline
%% $H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
%%   \cred{\sigma}$   $\rightsquigarrow
%% \rho, S, H,  e_2, \cred{\sigma}$} & \sc{if-false} \\
%% \hline
%% $H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
%%   (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma}$ $\rightsquigarrow
%% \rho', (\rho, x, (\SIF~x~e_1~e_2),  \cred{\sigma})\!:\!S, H, s,
%% \cred{\clazy\sigma}$}
%% &
%% \sc{if-clo} \\
%% \hline
%% {$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho,
%%   (\rho', z, e, \cred{\sigma'})\!:\!S, H, (\SRETURN~x), \cred{\sigma}$  $\rightsquigarrow \rho', S, H[\rho'(z) := v], e,
%%   \cred{\sigma'}$} &
%% \sc{return-whnf}\\
%% \hline
%% $H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
%%   (\psi:\SRETURN~x), \cred{\sigma}$  $
%%   \rightsquigarrow$
%% $\rho',~ (\rho, x, (\SRETURN~x), \cred{\sigma})\!:\!S, H,  s,
%%   \cred{\sigma}$} &
%% \sc{return-clo} \\
%% \hline
%% \end{tabular}
%% \caption{Minefield semantics.\label{fig:minefield-semantics}}
%% \end{center}
%% \vskip -5mm
%% \end{figure*}

@@

\hline
\makecell[t]{$\cred{GC(\rho_1, S, H_1, (\CAR~x), \sigma) = (\rho, S, H)}$,} $H(\rho(x)) \mbox{ is } \bot$ & $\rho, S,
  H, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$   &
\sc{car-clo-bang} 
\\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, H_1, (\CAR~x), \sigma) = (\rho, S, H)}$,} $H(\rho(x))
\mbox{ is } \langle s, \rho'\rangle$ & $\rho, S,
  H, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, x,
  (\CAR~x), \cred{\sigma})\!:\!S, H, s, \cred{(\clazy \cup \acar)\sigma }$         &
\sc{car-clo}\\
\hline
\end{tabular}
\end{center}
\end{figure*}

To set up the minefield semantics, we follow these steps:
\begin{enumerate}
\item   We   enrich    the   abstract   machine   state
  $\rho,S,H,e$   to  $\rho,S,H,e,\sigma$. 
We  call such a state  a \emph{minefield state}.
Here $\sigma$  is  the  ``dynamic'' demand  on  the  expression
  $e$, i.e., it arises
  from the  actual sequence of function  calls that led
  to  the evaluation  of  $e$.  %% The  0-CFA demand  used
  %% during    liveness     analysis    is     a    static
  %% over-approximation   of   $\sigma$   that   considers
  %% \emph{all  possible calls}  that  could  lead to  the
  %% evaluation of  $e$.
  The  information in  a suspended
  evaluation  context  on  the   stack  $S$  is  also
  similarly augmented with its  demand.  Thus a stacked
  entry now takes  the form $(\rho, \ell,  e, \sigma)$.  %% A
  %% modification  of  the   small  step  semantics  which
  %% carries the extra information  is shown in the bottom
  %% table of Figure~\ref{fig:minefield-semantics}.
\item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness  based  garbage
  collection that returns $(\rho', S', \heap')$, where the changes in
  $\rho, S$ and  $\heap$ are due
  to certain references being replaced by $\bot$.  This represents the
  act of  not retaining  (effectively garbage collecting)  the objects
  pointed to by these references  during an actual garbage collection.
  This is done in the following way:
  \begin{enumerate}
  \item Given the current  evaluation context $\rho, S,
    \heap,   e,  \sigma$   and   the  demand   transformers
    $\Lfonly$, we construct a liveness environment
    \Lv.  %%  If  $e$  is  an expression,  then  $  \Lv  =
    %% \mathcal{L}(e, \sigma, \Lfonly)$, else if $e$ is an
    %% application, then $ \Lv = ref(e, \sigma, \Lfonly)$.
    We  similarly  define the  liveness environment
    for each  of the  suspended evaluation  contexts in
    $S$, giving  a stack of liveness  environments that
    we shall denote $\mathsf{SL}$.
  \item \emph{Garbage collection  of objects pointed by
    root   set:}  For   each   $x  \in   domain(\rho)$,
    $\rho'(x)=\bot$     iff      $x.\epsilon     \notin
    \Lv$.    Similarly,    for   each    stack    entry
    $(\rho,\_,\_,\_)$  in   $S$  with  $\Lv'$   as  the
    corresponding      liveness     environment      in
    $\mathsf{SL}$, and  for each $x  \in domain(\rho)$,
    $\rho'(x)=\bot$ iff $x.\epsilon \notin \Lv'$.
  \item  \emph{Garbage  collection  of objects  in  the
    heap:} \warning{ For each location  $\ell$, $\heap'(\ell) = \bot$
    iff  there   is  no  $x$  in   either  the  current
    environment  or  one  of the  stacked  environments
    $\rho$, such  that for some  $\alpha$ corresponding
    to a  \warning{closure-free path}  in the heap  starting from
    $x$,  $\heap[x.\alpha]   =  \ell$  and   $x.\alpha  \in
    \Lv$}. Here \Lv\ is the liveness of the variables in
    the current environment or the stacked environment.
  \end{enumerate}
\end{enumerate}
Starting from the initial state,
%%  $([\;]_\rho,([\;]_\rho,
%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{H}, (\mainpgm)$,
every transition  is preceded  by a  garbage collection
using $GC$.  The details of the transition for the {\sc
  car-clo}  rule   is  shown   in  the  top   table  of
Figure~\ref{fig:minefield-semantics}.  If  this  or  an
earlier  call to  $GC$  results  in $H(\rho(x))$  being
bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
enters the  \bang\ state {\sc  car-clo-bang}. Otherwise
the transition is the same as the earlier {\sc car-clo}
rule.

>>>>

\begin{center}\footnotesize
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\
\hline
\hline 
          & $\rho,\, (\rho', \ell, e)\!:\!S,\, \heap,\, \kappa
  \rightsquigarrow \rho',\, S,\, \heap[\ell := \kappa],\, e$    &  \sc{const}
\\
\hline
          & {$\rho, \,(\rho', \ell, e)\!:\!S,\, \heap,\, (\CONS~x~y)
\rightsquigarrow
$  $\rho',\, S,\, \heap[\ell := (\rho(x), \rho(y))],\, e$}     &  \sc{cons} \\
\hline

$\heap(\rho(x)) \mbox{ is } (v, d)$ & $\rho,\, (\rho', \ell, e)\!:\!S,\, \heap,\,
(\CAR~x)  \rightsquigarrow \rho',\, S,\, \heap[\ell := v],\, e$      &
\sc{car-select} \\
\hline

$\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$ & $\rho,\, S,\,  \heap,\,
(\CAR~x)  \rightsquigarrow \rho', \,(\rho, addr(\langle s, \rho'\rangle), (\CAR~x))\!:\!S,\, \heap,\, s$      &
\sc{car-1-clo} \\
\hline

$\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho,\, S,\, \heap,\, (\CAR~x)
\rightsquigarrow
\rho',\, (\rho, \rho(x), (\CAR~x))\!:\!S, \,\heap,\, s$      &
\sc{car-clo}
\\
\hline

$\heap(\rho(x)), \heap(\rho(y)) \in \mathbb{N}$
 & {$\rho,\, (\rho', \ell, e)\!:\!S,\, \heap,\, (+~x~y)  \rightsquigarrow$
$\rho', \,S,\, \heap[\ell := \heap(\rho'(x)) + \heap(\rho'(y))],\, e$}      &
\sc{prim-add} \\
\hline

$\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho,\, S,\, \heap,\, (+~x~y)
\rightsquigarrow
\rho',\, (\rho, \rho(x), (+~x~y))\!:\!S,\, \heap,\, s$      &
\sc{prim-1-clo} \\
\hline
$\heap(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & $\rho,\, S,\, \heap,\, (+~x~y)
\rightsquigarrow
\rho', (\rho,\, \rho(y),\, (+~x~y))\!:\!S,\, \heap,\, s$      &
\sc{prim-2-clo} \\
\hline
{$\mathit{f}~\mbox{defined as}$
$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho,\, S,\, \heap,\,
(f~\myvec{x})  \rightsquigarrow
[\myvec{y} \mapsto \rho(\myvec{x})],\, S,\, \heap,\, e_{\mathit{f}}$      &
\sc{funcall} \\
\hline
$\ell$ is a new location& {$\rho,\, S,\, \heap,\, (\LET~x\leftarrow s~\IN~e)
  \rightsquigarrow$
$\rho\oplus[x \mapsto \ell],\, S,\, \heap[\ell \mapsto \langle s,
    \lfloor\rho\rfloor_{FV(s)} %% \oplus [x \mapsto  \ell] -- NO LAZY LETS
    \rangle],\, e$} &
\sc{let} \\
\hline
$\heap(\rho(x)) \ne 0$ & $\rho,\, S,\, \heap,\, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho,\, S,\, \heap,\,  e_1$ & \sc{if-true} \\
\hline
$\heap(\rho(x)) = 0$ & $\rho,\, S,\, \heap,\, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho,\, S, \,\heap, \, e_2$ & \sc{if-false} \\
\hline
$\heap(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho,\, S,\, \heap,\,
  (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho',\, (\rho, \rho(x), (\SIF~x~e_1~e_2))\!:\!S,\, \heap, \, s$}
&
\sc{if-clo} \\
\hline
{$\heap(\rho(x))~\mbox{is in WHNF with value}~v$}& $\rho,\, (\rho', \ell,
e)\!:\!S,\, \heap,\,
(\SRETURN~x)  \rightsquigarrow \rho',\, S,\, \heap[\ell := v],\, e$ &
\sc{return-whnf}\\
\hline
$\heap(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho,\, S,\, \heap,\, (\SRETURN~x)
  \rightsquigarrow$
$\rho',\, (\rho, \rho(x), (\SRETURN~x))\!:\!S,\, \heap,\,  s$} &
\sc{return-clo} \\
\hline
\end{tabular}
\caption{The small-step semantics for the language. \label{fig:lang-semantics}}
\end{center}
%}\comment
\end{figure*}

@@

The  initial  state  of  the  transition  system  described  above  is
$([\;]_\rho,\,   ([\;]_\rho,   \,   \rho(\ans),\,   (\print~\ans)):[\;]_{S},
[\;]_{\heap},\, (\mainpgm))$,  in which  $[\;]_\rho$ and $[\;]_{\heap}$  are the
empty environment and  the empty heap. The  initial stack consists
of  a single  continuation frame  in  which \ans\  is a  distinguished
variable  that  would   eventually  be  updated  with   the  value  of
(\mainpgm).  In  addition, \print\ is  a function modeling  a printing
mechanism---a   standard   run-time   support  assumption   for   lazy
languages---that prints  the value of  (\mainpgm).  The operator $:$
pushes elements on top of the stack.

@@

%% \hline
%% \makecell[t]{$GC(\rho_1, S, \heap_1, \sigma) = (\rho, S, \heap)$,}\\ $\heap(\rho(x)) \mbox{ is } \bot$ & $\rho, S,
%%   \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$   &
%% \sc{car-clo-bang}
%% \\
%% \hline
%% \makecell[t]{$GC(\rho_1, S, \heap_1, \Lv, \sigma) = (\rho, S, \heap)$,}\\ $\heap(\rho(x))
%% \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S,
%%   \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, x,
%%   (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$         &
%% \sc{car-clo}\\
%% \hline
%% \end{tabular}

@@

%%           &\makecell{ $\rho, (\rho', x, e, \cred{\sigma'})\!:\!S,
%%   \heap, \kappa, \cred{\sigma}$  $\rightsquigarrow \rho', S, \heap[\rho'(x) :=
%%     \kappa], e, \cred{\sigma'}$ }   &  \sc{const}
%% \\
%% \hline
%%           & \makecell{$\rho, (\rho', z, e, \cred{\sigma'})\!:\!S, \heap, (\CONS~x~y), \cred{\sigma}$  $\rightsquigarrow
%%   \rho', S, \heap[\rho'(z) := (\rho(x),\rho(y))], e, \cred{\sigma'}$}     &  \sc{cons} \\
%% \hline
%% $\heap(\rho(x)) \mbox{ is } (\ell_1, \ell_2)$ & \makecell{$\rho, (\rho', z, e,
%%   \cred{\sigma'} )\!:\!S, \heap, (\CAR~x), \cred{\sigma}$  $
%%   \rightsquigarrow \rho', S, \heap[\rho'(z) := \heap(\ell_1)], e, \cred{\sigma'}$}      &
%% \sc{car-whnf} \\
%% \hline
%% $\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
%%   \heap, (\CAR~x), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
%%   (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$}      &
%% \sc{car-clo}
%% \\

%% \hline
%% $\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$ & $\rho,\, S,\,  \heap,\,
%% (\CAR~x), \cred{\sigma} \rightsquigarrow \rho', \,(\rho, addr(\langle
%% s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,\, \heap,\, s, \, \cred{\sigma}$      &
%% \sc{car-1-clo} \\

%% \hline
%% $\heap(\rho(x)), \heap(\rho(y)) \in \mathbb{N}$
%%  & \makecell{$\rho, (\rho', z, e, \cred{\sigma'})\!:\!S, \heap,
%%   (+~x~y), \cred{\sigma}$   $\rightsquigarrow \rho', S, \heap[\rho'(z)
%%     := \heap(\rho'(x)) + \heap(\rho'(y))], e, \cred{\sigma'}$}      &
%% \sc{prim-whnf} \\
%% \hline
%% $\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S,
%%   \heap, (+~x~y), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
%%   (+~x~y), \cred{\sigma})\!:\!S, \heap, s, \cred{\clazy\sigma}$}      &
%% \sc{prim-1-clo} \\
%% \hline
%% $\heap(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho,
%%   S, \heap, (+~x~y), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, y,
%%   (+~x~y), \cred{\sigma})\!:\!S, \heap, s, \cred{\clazy\sigma}$}      &
%% \sc{prim-2-clo} \\
%% \hline
%% {$\mathit{g}~\mbox{defined as}$
%% $~(\DEFINE~(g~\myvec{y})~e_{\mathit{g}})$}  & \makecell{$\rho, S, \heap,
%%   (g~\myvec{x}), \cred{\sigma}$  $\rightsquigarrow [\myvec{y} \mapsto
%%     \rho(\myvec{x})], S, \heap, e_{\mathit{g}}, \cred{\sigma}$}      &
%% \sc{funcall} \\
%% \hline
%% $\ell$ is a new location& \makecell{$\rho, S, \heap, (\LET~x\leftarrow
%%   s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, \heap[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
%%   \ell]\rangle], e, \cred{\sigma}$} &
%% \sc{let} \\ 
%% \hline
%% $\heap(\rho(x)) \ne 0$ & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
%%   \cred{\sigma}$  $\rightsquigarrow \rho, S, \heap,  e_1, \cred{\sigma}$} & \sc{if-true} \\
%% \hline
%% $\heap(\rho(x)) = 0$ & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
%%   \cred{\sigma}$   $\rightsquigarrow
%% \rho, S, \heap,  e_2, \cred{\sigma}$} & \sc{if-false} \\
%% \hline
%% $\heap(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, \heap,
%%   (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma}$ $\rightsquigarrow
%% \rho', (\rho, x, (\SIF~x~e_1~e_2),  \cred{\sigma})\!:\!S, \heap, s,
%% \cred{\clazy\sigma}$}
%% &
%% \sc{if-clo} \\
%% \hline
%% {$\heap(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho,
%%   (\rho', z, e, \cred{\sigma'})\!:\!S, \heap, (\SRETURN~x), \cred{\sigma}$  $\rightsquigarrow \rho', S, \heap[\rho'(z) := v], e,
%%   \cred{\sigma'}$} &
%% \sc{return-whnf}\\
%% \hline
%% $\heap(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, \heap,
%%   (\psi:\SRETURN~x), \cred{\sigma}$  $
%%   \rightsquigarrow$
%% $\rho',~ (\rho, x, (\SRETURN~x), \cred{\sigma})\!:\!S, \heap,  s,
%%   \cred{\sigma}$} &
%% \sc{return-clo} \\
%% \hline
%% \end{tabular}
%% \caption{Minefield semantics.\label{fig:minefield-semantics}}
%% \end{center}
%% \vskip -5mm
%% \end{figure*}

@@

\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$,} $\heap(\rho(x)) \mbox{ is } \bot$ & $\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$   &
\sc{car-clo-bang} 
\\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$,} $\heap(\rho(x))
\mbox{ is } \langle s, \rho'\rangle$ & $\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, x,
  (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$         &
\sc{car-clo}\\
\hline
\end{tabular}
\end{center}
\end{figure*}

To set up the minefield semantics, we follow these steps:
\begin{enumerate}
\item   We   enrich    the   abstract   machine   state
  $\rho,S,\heap,e$   to  $\rho,S,\heap,e,\sigma$. 
We  call such a state  a \emph{minefield state}.
Here $\sigma$  is  the  ``dynamic'' demand  on  the  expression
  $e$, i.e., it arises
  from the  actual sequence of function  calls that led
  to  the evaluation  of  $e$.  %% The  0-CFA demand  used
  %% during    liveness     analysis    is     a    static
  %% over-approximation   of   $\sigma$   that   considers
  %% \emph{all  possible calls}  that  could  lead to  the
  %% evaluation of  $e$.
  The  information in  a suspended
  evaluation  context  on  the   stack  $S$  is  also
  similarly augmented with its  demand.  Thus a stacked
  entry now takes  the form $(\rho, \ell,  e, \sigma)$.  %% A
  %% modification  of  the   small  step  semantics  which
  %% carries the extra information  is shown in the bottom
  %% table of Figure~\ref{fig:minefield-semantics}.
\item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness  based  garbage
  collection that returns $(\rho', S', \heap')$, where the changes in
  $\rho, S$ and  $\heap$ are due
  to certain references being replaced by $\bot$.  This represents the
  act of  not retaining  (effectively garbage collecting)  the objects
  pointed to by these references  during an actual garbage collection.
  This is done in the following way:
  \begin{enumerate}
  \item Given the current  evaluation context $\rho, S,
    \heap,   e,  \sigma$   and   the  demand   transformers
    $\Lfonly$, we construct a liveness environment
    \Lv.  %%  If  $e$  is  an expression,  then  $  \Lv  =
    %% \mathcal{L}(e, \sigma, \Lfonly)$, else if $e$ is an
    %% application, then $ \Lv = ref(e, \sigma, \Lfonly)$.
    We  similarly  define the  liveness environment
    for each  of the  suspended evaluation  contexts in
    $S$, giving  a stack of liveness  environments that
    we shall denote $\mathsf{SL}$.
  \item \emph{Garbage collection  of objects pointed by
    root   set:}  For   each   $x  \in   domain(\rho)$,
    $\rho'(x)=\bot$     iff      $x.\epsilon     \notin
    \Lv$.    Similarly,    for   each    stack    entry
    $(\rho,\_,\_,\_)$  in   $S$  with  $\Lv'$   as  the
    corresponding      liveness     environment      in
    $\mathsf{SL}$, and  for each $x  \in domain(\rho)$,
    $\rho'(x)=\bot$ iff $x.\epsilon \notin \Lv'$.
  \item  \emph{Garbage  collection  of objects  in  the
    heap:} \warning{ For each location  $\ell$, $\heap'(\ell) = \bot$
    iff  there   is  no  $x$  in   either  the  current
    environment  or  one  of the  stacked  environments
    $\rho$, such  that for some  $\alpha$ corresponding
    to a  \warning{closure-free path}  in the heap  starting from
    $x$,  $\heap[x.\alpha]   =  \ell$  and   $x.\alpha  \in
    \Lv$}. Here \Lv\ is the liveness of the variables in
    the current environment or the stacked environment.
  \end{enumerate}
\end{enumerate}
Starting from the initial state,
%%  $([\;]_\rho,([\;]_\rho,
%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{\heap}, (\mainpgm)$,
every transition  is preceded  by a  garbage collection
using $GC$.  The details of the transition for the {\sc
  car-clo}  rule   is  shown   in  the  top   table  of
Figure~\ref{fig:minefield-semantics}.  If  this  or  an
earlier  call to  $GC$  results  in $\heap(\rho(x))$  being
bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
enters the  \bang\ state {\sc  car-clo-bang}. Otherwise
the transition is the same as the earlier {\sc car-clo}
rule.
